---
title: "Prediction Project Assignment"
course: "Practical Machine Learning" 
author: "KB"
date: "Saturday, May 28, 2016"
output: html_document
---

###Summary

Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement - a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. 
In this project, our goal will be to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. They were asked to perform barbell lifts correctly and incorrectly in 5 different ways. 
More information is available from the website here:
<http://groupware.les.inf.puc-rio.br/har> (see the section on the Weight Lifting Exercise Dataset).

###Goal

The goal is to predict the manner in which they did the exercise.

###Exploratory Data Analysis

The two sets of data used in this assignments are :

- Training data available at <https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv> (csv format)
- Test data available at <https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv> (csv format)

The data for this project come from this source: <http://groupware.les.inf.puc-rio.br/har>. 

####Loading Data

```{r, echo=FALSE, message=F, warning=F}
#Load libraries
suppressMessages(library(caret, warn.conflicts = FALSE, quietly=TRUE))
suppressMessages(library(AppliedPredictiveModeling, warn.conflicts = FALSE, quietly=TRUE))
suppressMessages(library(randomForest, warn.conflicts = FALSE, quietly=TRUE))
suppressMessages(library(rattle, warn.conflicts = FALSE, quietly=TRUE))
suppressMessages(library(ggplot2, warn.conflicts = FALSE, quietly=TRUE))
```

```{r}
#URLs of datasets
URLTrain <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
URLtest <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"

#Downloading sets of data
if (!file.exists("pml-training.csv")) {
  download.file(URLTrain, destfile = "pml-training.csv")
}
if (!file.exists("pml-testing.csv")) {
  download.file(URLtest, destfile = "pml-testing.csv")
}
TrainSet <- read.csv("pml-training.csv", sep = ",", na.strings = c("", "NA"))
TestSet <- read.csv("pml-testing.csv", sep = ",", na.strings = c("", "NA"))
```

Training dataset consists of 19622 rows and 160 columns
Testing dataset consists of 20 rows and 160 columns

```{r, echo=FALSE}
dim(TrainSet)
dim(TestSet)
```

#### Data partitioning

Since the testing dataset will only be used at the end of the project, to answer questions about our modeling process, the training dataset is partitioned in two sets:
- a training set of 70 % of the data 
- a testing set of 30 % of the data
We will partition the data based on the "classe" variable. The classe variable consist of 5 values : A, B, C, D and E.

```{r}
#Partitioning training data
set.seed(3433)
inTrain  <- createDataPartition(TrainSet$classe, p=0.7, list=FALSE)
training <- TrainSet[inTrain, ]
testing  <- TrainSet[-inTrain, ]
dim(training)
dim(testing)
```

####Cleaning Data

To build our model and predict classes, we need to remove data columns that contains mostly NAs and work with the ones with non-zero. After finding those columns, we narrow down the number of columns to 60.

```{r}
AllNA    <- sapply(training, function(x) mean(is.na(x))) > 0.95
training <- training[, AllNA==FALSE]
testing  <- testing[, AllNA==FALSE]
```

Since the first 7 columns contains the person's identification and timestamps, we proceed to remove the first 7 columns.

```{r}
training <- training[, -(1:7)]
testing  <- testing[, -(1:7)]

#Datasets must have same columns
newcolnames <- colnames(training)
testing <- testing[newcolnames]

dim(training)
```

Both datasets, training and testing must have the same data types. In order to do that, we coerce the data.

```{r}
for (i in 1:length(testing) ) {
    for(j in 1:length(training)) {
        if( length( grep(names(training[i]), names(testing)[j]) ) == 1)  {
            class(testing[j]) <- class(training[i])
        }      
    }      
}
```

####Column correlation to Classe

We look for columns that are highly correlated to "classe" variable, considering 90 % as high correlation rate.

```{r}
outcome = which(names(training) == "classe")
predictors = findCorrelation(abs(cor(training[,-outcome])),0.90)
colnames(training[predictors])
#training = training[,-predictors]
```

###Prediction Model building

I will build three different model algorithms and apply them to the training and testing datasets we built from the original training set. Also, we will do model comparison using the confusion matrix function, which will give us information about how well the model's did on new data sets.

Cross validation will be performed for each model (3 models).

```{r}
tC <- trainControl(method='cv', number = 3)
```

####Model 1: Generalized Boosted Model (gbm package)

```{r, message=F, warning=F, results='hide'}
ModelGBM <- train(classe ~ ., data=training, trControl=tC, method='gbm')
predictGBM <- predict(ModelGBM, newdata=testing)
cmGBM <- confusionMatrix(predictGBM, testing$classe)
```

####Model 2: Decision trees (rpart package)
```{r, message=F, warning=F}
ModelDT <- train(classe ~ ., data=training, method='rpart', trControl=tC)
#fancyRpartPlot(ModelDT)
predictDT <- predict(ModelDT, newdata=testing)
cmDT <- confusionMatrix(predictDT, testing$classe)
```
 
####Model 3: Random forest decision trees (rf)

```{r, message=F, warning=F}
ModelRF <- train(classe ~ ., data=training, trControl=tC, method='rf', ntree=100)
predictRF <- predict(ModelRF, newdata=testing)
cmRF <- confusionMatrix(predictRF, testing$classe)
```

```{r, message=F, warning=F}
ModelRF2 <- rpart(classe ~ ., data=training, method="class")
fancyRpartPlot(ModelRF2)
```

####Cross Validation

```{r}
AccuracyValues <- data.frame( Model = c('GBM', 'DT', 'RF'), Accuracy = rbind(cmGBM$overall[1], cmDT$overall[1], cmRF$overall[1]))
print(AccuracyValues)
```

Based on the accuracy results, we identify Decision Tree Model as the least accurate, and Random Forest Model slightly outperforms GBM Model.

```{r}
RFcm = randomForest(classe ~ ., data = training, ntree=500, importance = TRUE)
#varImp(RFcm, scale = FALSE)
head(importance(RFcm))
varImpPlot(RFcm,type=2)
```

We can identify our best predictors as : roll_belt, yaw_belt, pitch_forearm, magnet_dumbbell_x, pitch_belt.

###Prediction

Now that we found that Random Forest is the model we will use to predict a classe for each of the 20 observations contained in the testing dataset from file ('pml-testing.csv'), we do the prediction with the testing data.

```{r}
predictTesting <- predict(ModelRF, newdata=TestSet)
QuizResults <- data.frame(problem_id=TestSet$problem_id, predicted=predictTesting)
print(QuizResults)
```

```{r , echo=FALSE}
# Write the results to a text file for submission
pml_write_files = function(x){
    n = length(x)
    for(i in 1:n){
        filename = paste0("problem_id_",i,".txt")
        write.table(x[i],file=filename,quote=FALSE,row.names=FALSE,col.names=FALSE)
    }
}
pml_write_files(predictTesting)
```

###Conclusion

Based on the data was available at the time, Random Forest Model with cross validation was implemented due to its 99.2 % accuracy, to predict the 20 sample observations given as testing.
One thing to note, is lots of missing data in both datasets. We proceeded to remove some of those columns with NAs.